preciseTAD: A machine learning framework for precise TAD boundary prediction at base-level resolution

### 2020-08-26 Manuscript/manuscript.Rmd
completed todo items

We assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture, including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* and *plotHeatmap* tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy] [??? Correct, change to new pipeline]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

TAD boundaries are relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35%/55% of *preciseTAD*-predicted domain boundaries using models trained on Arrowhead/Peakachu data, respectively, were conserved between GM12878 and K562 cell lines. This is in contrast to only 26% and 55% for Arrowhead and Peakachu boundaries themselves (Figure 6B). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

@@

**Supplementary Figure 9.** *preciseTAD*-predicted TAD boundaries using Arrowhead are nested within *preciseTAD*-predicted chromatin loop boundaries using Peakachu. Venn diagrams illustrating the agreement between boundaries predicted using Arrowhead- and Peakachu-trained models. Domain boundaries were flanked by resolution (5 kb for models using Arrowhead and 10 kb for Peakachu).

>>>>

We assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture, including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix*, *plotProfile*, and *plotHeatmap* tools offered by deepTools (version 2.0) [@ramirez2016deeptools] <!--via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy]--> [??? Correct, change to new pipeline; !!!removed galaxy usage]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

TAD boundaries are relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 46%/63% of *preciseTAD*-predicted domain boundaries using models trained on Arrowhead/Peakachu data, respectively, were conserved between GM12878 and K562 cell lines. This is in contrast to only 26% and 63% for Arrowhead and Peakachu boundaries themselves (Figure 6B). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 86.8% and 92.6% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

@@

**Supplementary Figure 9.** *preciseTAD*-predicted TAD boundaries using Arrowhead are nested within *preciseTAD*-predicted chromatin loop boundaries using Peakachu. Venn diagrams illustrating the agreement between boundaries predicted using Arrowhead- and Peakachu-trained models for (A) GM12878 and (B) K562. Domain boundaries were flanked by resolution (5 kb for models using Arrowhead and 10 kb for Peakachu).
